Modularizing Street Map Generation \\ \vspace{8pt} \large An Adventure in Tensors

### 2013-12-12 paper/paper.tex
rewrote assessment and conclusion

\section{Mathematical Background}\label{sec:math}
In this section we will introduce the mathematical concepts necessary to
understand our algorithm. A tensor $t$ is a geometric object represented by a
2x2 symmetric, traceless matrix:
\[
    \sqmat{\cos{2\theta}}{\sin{2\theta}}{\sin{2\theta}}{-\cos{2\theta}}
    = \sqmat{a}{b}{b}{-a}
\]
where $R$ is the magnitude and $\theta$ is the direction, and
$R\geq0 \textrm{ and } \theta\in[0,2\pi)$. Since the matrix is symmetric and
traceless, it is only necessary to store two values, $a$ and $b$ to encode
the tensor.

@@

\subsection{Eigenvectors}
The eigenvalues of a tensor $t$ represented by the matrix $\sqmat{a}{b}{b}{-a}$
are:
\begin{align}
    \lambda_1 &= \sqrt{a^2+b^2} \\
    \lambda_2 &= -\lambda_1 \\
\end{align}

@@

\subsection{Streamline Tracing}
Mathematically, the basic streamline tracing algorithm is fairly simple,
however, in order to produce a set of streamlines that approximate a road map,
there are several important cases to consider. In order to trace the
streamlines, we are forced to use a heuristic, since it isn’t possible to trace
a continuous line over a constantly-changing tensor field. In order to
approximate this, we choose a step size $d_\textrm{step}$. Beginning at a seed
point, which is chosen using the methods detailed in Section \ref{seedpoints},
we calculate the eigenvector at that point, extend a line in that direction
with length $d_\textrm{step}$, and check if we have reached an end condition.
This process is repeated on the line until one of three end conditions is
reached:
\begin{enumerate}
    \item The point is outside the bounds of the map area
    \item The point is within a certain distance of an existing streamline.
    \item The point is part of a cycle
\end{enumerate}

@@

\subsection{Expressiveness}
As our work was not a complete reimplementation of the system described in
\cite{chen}

@@

\subsection{Modularity \& Extensibility}

In an attempt to compare this with the original work, we consulted the header
files available with Chen’s code. The system has significantly more
functionality than ours, but we had some difficulty discerning which modules
were parallel to ours. The \texttt{tensoranalysis} contains methods for tensor
line tracing, and \texttt{ClosedStreamlineTracing} seems to be related,
although we did not find documentation on the difference between open and
closed tracing. \texttt{SeedPointCtrlDlg} is a dialogue probably related to the
OpenGL framework, but didn’t have the actual calculations of seed points. At
the least, our implementation has a more direct pipeline-to-module correlation,
which will be beneficial for future contributors who wish to build off of our
work.

\subsection{Natural Expression of Underlying Mathematics}
Overall, this seemed to be successful. In general, functional languages are
well suited to express mathematics. Our decision to implement this program in
Haskell meant that that we were able to take advantage of this natural
proclivity towards math in order to make our code easy to understand.

An example of this can be seen in the calculation of eigenvectors. We derived
the formula for the eigenvector of a symmetric, traceless tensor
(see Section~\ref{sec:math}) and implemented it in the functions in
Figure~\ref{fig:evecs}.

In [Chen], there is a 75-line function,
\texttt{cal\_eigenvecs\_onevert\_quad}, to do approximately the same thing. There
is a fair amount of documentation in the body of the function, explaining what
each line does, but there also seem to be some side effects relating to global
variables and large data structures that the authors found difficult to follow,
mostly by the nature of the language.  Generally speaking, we found that the
mathematics was significantly easier to express and review in a Haskell
implementation.

\section{Conclusion}
In this paper we have presented the underlying mathematics and functional
implementation of the work described in \cite{chen}, with the goals being that a
functional implementation would easily express the mathematics and provide
modularity that the imperative implementation provide in \cite{chen} could not.

>>>>

\section{Mathematical Background}\label{sec:math}
In this section we will introduce the mathematical concepts necessary to
understand our algorithm. A tensor $t$ is a geometric object represented by a
$2\times2$ symmetric, traceless matrix:
\[
    \sqmat{\cos{2\theta}}{\sin{2\theta}}{\sin{2\theta}}{-\cos{2\theta}}
    = \sqmat{a}{b}{b}{-a}
\]
where $R$ is the magnitude and $\theta$ is the direction, and
$R\geq0 \textrm{ and } \theta\in[0,2\pi)$. Since the matrix is symmetric and
traceless, it is only necessary to store two values, $a$ and $b$ to encode
the tensor.

@@

\subsection{Eigenvectors}\label{sec:eigenvectors}
The eigenvalues of a tensor $t$ represented by the matrix $\sqmat{a}{b}{b}{-a}$
are:
\begin{align}
    \lambda_1 &= \sqrt{a^2+b^2} \\
    \lambda_2 &= -\lambda_1 \\
\end{align}

@@

\subsection{Streamline Tracing}
Mathematically, the basic streamline tracing algorithm is fairly simple,
however, in order to produce a set of streamlines that approximate a road map,
there are several important cases to consider. In order to trace the
streamlines, we are forced to use a heuristic, since it isn’t possible to trace
a continuous line over a constantly-changing tensor field. In order to
approximate this, we choose a step size $d_\textrm{step}$. Beginning at a seed
point, which is chosen using the methods detailed in Section
\ref{sec:seedpoints}, we calculate the eigenvector at that point, extend a line
in that direction with length $d_\textrm{step}$, and check if we have reached
an end condition.  This process is repeated on the line until one of three end
conditions is reached:

\begin{enumerate}
    \item The point is outside the bounds of the map area
    \item The point is within a certain distance of an existing streamline.
    \item The point is part of a cycle
\end{enumerate}

@@

\subsection{Expressiveness}
As our work was not a complete reimplementation of the system described in
\cite{chen}, it is difficult to compare the relative expressiveness of our
Haskell code with the \cpp code of the original implementation. However, we
will present one comparison of functions from our respective implementations
with significantly overlapping functionality. Our Haskell code for calculating
the eigenvectors for a tensor can be found in Figure \ref{fig:evecs}. The
reader may note the close correspondence between the body of these functions
and the mathematical formulas used for calculating eigenvectors in Section
\ref{sec:eigenvectors}.  The 75-line function
\texttt{cal\_eigenvecs\_onevert\_quad} in Chen's implementation. While it is
documented, we found it difficult to determine the side effects it had. In a
language without mutable state or side effect, we found it significantly easier
to fully understand the behavior of functions.  In contrast, even well-written
code in \cpp or another language in the imperative paradigm can be difficult to
understand, because the possibility for side effect requires the reader to have
an understanding of any part of the system that the function can affect, which
in practice can make the effort to understand any small part of a large
codebase much greater.

@@

\section{Conclusion}
In this paper we have presented the underlying mathematics and functional
implementation of the work described in \cite{chen}, with the goals being that
a functional implementation would readily express the mathematics, provide
modularity that the imperative implementation provide in \cite{chen} could not.
We believe that, due to the focus on expressibility and modularity, assessed in
the previous section, that an individual looking to understand the pipeline
will find our code undestandable and approachable, and lend itself to be
easily extended. We hope that our work will motivate a greater use of
functional languages in future efforts in the research of new procedural
algorithms.
